Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-03-25 pkg/disttree/inst/draft/draft.Rnw
edited appendix A
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2078 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b33ee8556344dfc4124d758d78310a39b872f15a

# plot RMSE
plot_rmse <- function(simres, type = c("exp", "par"), ylim = NULL, legend = TRUE){

  if(type == "exp"){
    rmse <- NULL
    colnames <- NULL
    col <- NULL
    legendnames <- NULL
    if("av.rmse.exp.true.dt" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.dt"], simres$resmat[,"av.rmse.exp.obs.dt"])
      colnames <- c(colnames, "dt.true", "dt.obs")
      col <- c(col, pal["tree"])
      legendnames <- c(legendnames, "tree")
    }
    if("av.rmse.exp.true.df" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.df"], simres$resmat[,"av.rmse.exp.obs.df"])
      colnames <- c(colnames, "df.true", "df.obs")
      col <- c(col, pal["forest"])
      legendnames <- c(legendnames, "distributional forest")
    }
    if("av.rmse.exp.true.g" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.g"], simres$resmat[,"av.rmse.exp.obs.g"])
      colnames <- c(colnames, "g.true", "g.obs")
      col <- c(col, pal["gamlss"])
      legendnames <- c(legendnames, "gamlss")
    }
    if("av.rmse.exp.true.b" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.b"], simres$resmat[,"av.rmse.exp.obs.b"])
      colnames <- c(colnames, "b.true", "b.obs")
      col <- c(col, pal["bamlss"])
      legendnames <- c(legendnames, "bamlss")
    }
    if("av.rmse.exp.true.gb" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.gb"], simres$resmat[,"av.rmse.exp.obs.gb"])
      colnames <- c(colnames, "gb.true", "gb.obs")
      col <- c(col, pal["gamboostLSS"])
      legendnames <- c(legendnames, "gamboostLSS")
    }
    if("av.rmse.exp.true.rf" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.rf"], simres$resmat[,"av.rmse.exp.obs.rf"])
      colnames <- c(colnames, "rf.true", "rf.obs")
      col <- c(col, pal["randomForest"])
      legendnames <- c(legendnames, "randomForest")
    }
    if("av.rmse.exp.true.cf" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.exp.true.cf"], simres$resmat[,"av.rmse.exp.obs.cf"])
      colnames <- c(colnames, "cf.true","cf.obs")
      col <- c(col, pal["cforest"])
      legendnames <- c(legendnames, "cforest")
    }

    colnames(rmse) <- colnames
    if(is.null(ylim)) ylim <- c(min(na.omit(rmse)), max(na.omit(rmse)))

    plot(x = simres$x.axis, y = rmse[,1], type = "l", col = col[1], ylim = ylim,
         xlab = expression(kappa), ylab = "RMSE")
    lines(x = simres$x.axis, y = rmse[,2], type = "l", lty = 2, col = col[1])

    for(i in 1:(length(col)-1)){
      lines(x = simres$x.axis, y = rmse[,(2*i+1)], type = "l", col = col[i+1])
      lines(x = simres$x.axis, y = rmse[,(2*i+2)], type = "l", lty = 2, col = col[i+1])
    }
    if(legend) legend('topleft', legendnames, 
                      col = col, lty = 1, cex = 0.7)
  }

  if(type == "par"){

    rmse <- NULL
    colnames <- NULL
    col <- NULL
    legendnames <- NULL
    if("av.rmse.exp.true.dt" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.dt"], simres$resmat[,"av.rmse.sigma.dt"])
      colnames <- c(colnames, "dt.true", "dt.obs")
      col <- c(col, pal["tree"])
      legendnames <- c(legendnames, "tree")
    }
    if("av.rmse.exp.true.df" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.df"], simres$resmat[,"av.rmse.sigma.df"])
      colnames <- c(colnames, "df.true", "df.obs")
      col <- c(col, pal["forest"])
      legendnames <- c(legendnames, "distributional forest")
    }
    if("av.rmse.exp.true.g" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.g"], simres$resmat[,"av.rmse.sigma.g"])
      colnames <- c(colnames, "g.true", "g.obs")
      col <- c(col, pal["gamlss"])
      legendnames <- c(legendnames, "gamlss")
    }
    if("av.rmse.exp.true.b" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.b"], simres$resmat[,"av.rmse.sigma.b"])
      colnames <- c(colnames, "b.true", "b.obs")
      col <- c(col, pal["bamlss"])
      legendnames <- c(legendnames, "bamlss")
    }
    if("av.rmse.exp.true.gb" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.gb"], simres$resmat[,"av.rmse.sigma.gb"])
      colnames <- c(colnames, "gb.true", "gb.obs")
      col <- c(col, pal["gamboostLSS"])
      legendnames <- c(legendnames, "gamboostLSS")
    }
    if("av.rmse.exp.true.rf" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.rf"], simres$resmat[,"av.rmse.sigma.rf"])
      colnames <- c(colnames, "rf.true", "rf.obs")
      col <- c(col, pal["randomForest"])
      legendnames <- c(legendnames, "randomForest")
    }
    if("av.rmse.exp.true.cf" %in% colnames(simres$resmat)) {
      rmse <- cbind(rmse, simres$resmat[,"av.rmse.mu.cf"], simres$resmat[,"av.rmse.sigma.cf"])
      colnames <- c(colnames, "cf.true","cf.obs")
      col <- c(col, pal["cforest"])
      legendnames <- c(legendnames, "cforest")
    }

    colnames(rmse) <- colnames
    if(is.null(ylim)) ylim <- c(min(na.omit(rmse)), max(na.omit(rmse)))

    plot(x = simres$x.axis, y = rmse[,1], type = "l", col = col[1], ylim = ylim,
         xlab = expression(kappa), ylab = "RMSE")
    lines(x = simres$x.axis, y = rmse[,2], type = "l", lty = 2, col = col[1])

    for(i in 1:(length(col)-1)){
      lines(x = simres$x.axis, y = rmse[,(2*i+1)], type = "l", col = col[i+1])
      lines(x = simres$x.axis, y = rmse[,(2*i+2)], type = "l", lty = 2, col = col[i+1])
    }
    if(legend) legend('topleft', legendnames, 
                      col = col, lty = 1, cex = 0.7)
  }
}

# plot loglikelihood
plot_ll <- function(simres, ylim = NULL, legend = TRUE){

  ll <- NULL
  colnames <- NULL
  col <- NULL
  legendnames <- NULL
  if("av.rmse.exp.true.dt" %in% colnames(simres$resmat)) {
    ll <- cbind(ll, simres$resmat[,"av.loglik.dt"])
    colnames <- c(colnames, "dt.ll")
    col <- c(col, pal["tree"])
    legendnames <- c(legendnames, "tree")
  }
  if("av.rmse.exp.true.df" %in% colnames(simres$resmat)) {
    ll <- cbind(ll, simres$resmat[,"av.loglik.df"])
    colnames <- c(colnames, "df.ll")
    col <- c(col, pal["forest"])
    legendnames <- c(legendnames, "distributional forest")
  }
  if("av.rmse.exp.true.g" %in% colnames(simres$resmat)) {
    ll <- cbind(ll, simres$resmat[,"av.loglik.g"])
    colnames <- c(colnames, "g.true")
    col <- c(col, pal["gamlss"])
    legendnames <- c(legendnames, "gamlss")
  }
  if("av.rmse.exp.true.b" %in% colnames(simres$resmat)) {
    ll <- cbind(ll, simres$resmat[,"av.loglik.b"])
    colnames <- c(colnames, "b.true")
    col <- c(col, pal["bamlss"])
    legendnames <- c(legendnames, "bamlss")
  }
  if("av.rmse.exp.true.gb" %in% colnames(simres$resmat)) {
    ll <- cbind(ll, simres$resmat[,"av.loglik.gb"])
    colnames <- c(colnames, "gb.true")
    col <- c(col, pal["gamboostLSS"])
    legendnames <- c(legendnames, "gamboostLSS")
  }
  if("av.rmse.exp.true.rf" %in% colnames(simres$resmat)) {
    ll <- cbind(ll, simres$resmat[,"av.loglik.rf"])
    colnames <- c(colnames, "rf.true")
    col <- c(col, pal["randomForest"])
    legendnames <- c(legendnames, "randomForest")
  }
  if("av.rmse.exp.true.cf" %in% colnames(simres$resmat)) {
    ll <- cbind(ll, simres$resmat[,"av.loglik.cf"])
    colnames <- c(colnames, "cf.true")
    col <- c(col, pal["cforest"])
    legendnames <- c(legendnames, "cforest")
  }

  colnames(ll) <- colnames
  if(is.null(ylim)) ylim <- c(min(na.omit(ll)), max(na.omit(ll)))

  plot(x = simres$x.axis, y = ll[,1], type = "l", col = col[1], ylim = ylim,
       xlab = expression(kappa), ylab = "log-likelihood")

  for(i in 2:length(col)){
    lines(x = simres$x.axis, y = ll[,i], type = "l", col = col[i])
  }
  if(legend) legend('topleft', legendnames, 
                    col = col, lty = 1, cex = 0.7)
}

# plot crpse
plot_crps <- function(simres, ylim = NULL, legend = TRUE){

  crps <- NULL
  colnames <- NULL
  col <- NULL
  legendnames <- NULL
  if("av.rmse.exp.true.dt" %in% colnames(simres$resmat)) {
    crps <- cbind(crps, simres$resmat[,"av.crps.dt"])
    colnames <- c(colnames, "dt.crps")
    col <- c(col, pal["tree"])
    legendnames <- c(legendnames, "tree")
  }
  if("av.rmse.exp.true.df" %in% colnames(simres$resmat)) {
    crps <- cbind(crps, simres$resmat[,"av.crps.df"])
    colnames <- c(colnames, "df.crps")
    col <- c(col, pal["forest"])
    legendnames <- c(legendnames, "distributional forest")
  }
  if("av.rmse.exp.true.g" %in% colnames(simres$resmat)) {
    crps <- cbind(crps, simres$resmat[,"av.crps.g"])
    colnames <- c(colnames, "g.true")
    col <- c(col, pal["gamlss"])
    legendnames <- c(legendnames, "gamlss")
  }
  if("av.rmse.exp.true.b" %in% colnames(simres$resmat)) {
    crps <- cbind(crps, simres$resmat[,"av.crps.b"])
    colnames <- c(colnames, "b.true")
    col <- c(col, pal["bamlss"])
    legendnames <- c(legendnames, "bamlss")
  }
  if("av.rmse.exp.true.gb" %in% colnames(simres$resmat)) {
    crps <- cbind(crps, simres$resmat[,"av.crps.gb"])
    colnames <- c(colnames, "gb.true")
    col <- c(col, pal["gamboostLSS"])
    legendnames <- c(legendnames, "gamboostLSS")
  }
  if("av.rmse.exp.true.rf" %in% colnames(simres$resmat)) {
    crps <- cbind(crps, simres$resmat[,"av.crps.rf"])
    colnames <- c(colnames, "rf.true")
    col <- c(col, pal["randomForest"])
    legendnames <- c(legendnames, "randomForest")
  }
  if("av.rmse.exp.true.cf" %in% colnames(simres$resmat)) {
    crps <- cbind(crps, simres$resmat[,"av.crps.cf"])
    colnames <- c(colnames, "cf.true")
    col <- c(col, pal["cforest"])
    legendnames <- c(legendnames, "cforest")
  }

  colnames(crps) <- colnames
  if(is.null(ylim)) ylim <- c(min(na.omit(crps)), max(na.omit(crps)))

  plot(x = simres$x.axis, y = crps[,1], type = "l", col = col[1], ylim = ylim,
       xlab = expression(kappa), ylab = "CRPS")

  for(i in 2:length(col)){
    lines(x = simres$x.axis, y = crps[,i], type = "l", col = col[i])
  }
  if(legend) legend('topright', legendnames, 
                    col = col, lty = 1, cex = 1, bty = "n")
}
@

@@

After considering only one observation station up to now, a second step
evaluates and compares the competing methods on all 95~stations. As in the previous
section, all models are learned on the first 24~years and evaluated by the
average CRPS on the last 4~years. More specifically, the CRPS skill score
against the EMOS model is computed for the out-of-sample predictions at each
station and visualized by parallel coordinates plots with boxplots superimposed
in Figure~\ref{fig:all-crps}. Overall distributional forests have a slightly
higher improvement in CRPSS compared to the two GAMLSS which is best seen
by looking at the boxplots and also in the green line representating the results
for station Axams. The underlying parallel coordinates additionally
bring out that the prespecified GAMLSS sometimes performs rather differently
(sometimes better, sometimes worse) compared to the two data-driven models.
Finally, because some CRPSS become negative, it is also clear that the simple
EMOS performs better than the other models for a few of the stations.

@@

The tree algorithm used in the applications discussed in this paper 
is going to be explained following the ideas presented in 
\cite{Seibold+Zeileis+Hothorn:2017}. For notational simplicity, the testing 
and splitting procedure is described for the root node with observations 
$\{y_i\}_{i = 1,\ldots,n}$, $n \in \mathbb{N}$. In each child node the 
corresponding subset of observations depends on the assignment made 
by the foregoing split.

\begin{itemize}
\item A distributional model $\mathcal{D}(Y, \bm{\theta})$ is fit to the set of 
observations $\{y_i\}_{i = 1,\ldots,n}$ by estimating the distribution 
parameter $\bm{\hat \theta}$. This parameter can be a multidimensional 
vector $(\hat{\theta}_1, \ldots, \hat{\theta}_K)$, $K \in \mathbb{N}$, 
depending on the type of distribution. The estimation is done by 
maximizing the log-likelihood function $\ell(\bm{\theta}; Y)$, i.e.
$$
\hat{\theta} = \argmax_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
$$
which is equal to solving
$$
\sum_{i=1}^n s(\theta, y_i) = \sum_{i=1}^n \frac{\partial \ell}{\partial \theta}(\theta; y_i) = 0
$$
for $\theta$.

\item Calculate score values 
$$
s(\hat{\theta}, y_i) = 
\begin{pmatrix} 
s(\hat{\theta}, y_1) \\
s(\hat{\theta}, y_2) \\
\vdots \\
s(\hat{\theta}, y_n)
\end{pmatrix} =
\begin{pmatrix} 
s(\hat{\theta}, y_1)_1 & s(\hat{\theta}, y_1)_2 & \ldots & s(\hat{\theta}, y_1)_K\\
s(\hat{\theta}, y_2)_1 & s(\hat{\theta}, y_2)_2 & \ldots & s(\hat{\theta}, y_2)_K\\
\vdots & \vdots & \ddots & \vdots \\
s(\hat{\theta}, y_n)_1 & s(\hat{\theta}, y_n)_2 & \ldots & s(\hat{\theta}, y_n)_K
\end{pmatrix}
$$

\item Test for independence between each column of scores $\left(s(\hat{\theta}, y_i)_k\right)_{i=1,\ldots,n}$ for $k\in\{1,\ldots,K\}$, and split variable $Z_j \in \{Z_1, \ldots, Z_m\}$.  
\begin{align*}
H_0^{1,j}:  \left(s(\hat{\theta}, y_i)_1\right)_{i=1,\ldots,n} \qquad \bot \qquad Z_j \\
H_0^{2,j}:  \left(s(\hat{\theta}, y_i)_2\right)_{i=1,\ldots,n} \qquad \bot \qquad Z_j \\
\vdots \hspace{3cm}\\
H_0^{K,j}:  \left(s(\hat{\theta}, y_i)_K\right)_{i=1,\ldots,n} \qquad \bot \qquad Z_j 
\end{align*}

To test these hypotheses permutation testing with the multivariate linear statistic
$$
T_{j} = \sum_{i=1}^n g_j(z_{ji}) \cdot s(\hat{\theta}, y_i)
$$
is applied. The type of the transformation function $g_j$ depends on 
the type of the split variable $Z_j$. If $Z_j$ is numeric then $g_j$ 
is simply the identity function $g_j(z_{ji}) = z_{ji}$. If $Z_j$ is 
a categorical variable with $L$ categories then 
$g_j(z_{ji}) = e_L(z_{ji}) = (\I(z_{ji} = 1), \ldots, \I(z_{ji} = L))$ 
such that $g_j$ is a unit vector where the element corresponding to 
the value of $z_{ji}$ is $1$. Observations with missing values are 
excluded from the sums.

With the conditional expectation $\mu_{j}$ and the covariance 
$\Sigma_{j}$ as derived by \cite{Strasser+Weber:1999} the test 
statistic can be standardized. For a numeric split variable this 
results in the Pearson correlation coefficient
$$
c(t_{j},\mu_{j},\Sigma_{j}) = \left|\frac{t_{j} - \mu_{j}}{\sqrt{\Sigma_{j}}}\right|.
$$
Otherwise the standardized form is
$$
c(t_{j},\mu_{jk},\Sigma_{j}) = \max_{l=1,\ldots,L}\left|\frac{(t_{j} - \mu_{j})_l}{\sqrt{(\Sigma_{j})_{ll}}}\right|.
$$
The smaller the p-value corresponding to the standardized test 
statistic $c(t_{j},\mu_{j},\Sigma_{j})$ is the stronger the discrepancy 
from the assumption of independence between the scores and the split 
variable $Z_j$.

\item After Bonferroni adjusting the p-values check whether any of 
the resulting p-values is beneath the selected significance level. 
If so, choose the partitioning variable $Z_{j^\ast}$ with the lowest 
p-value to the scores corresponding to the distribution parameters 
relevant for the split.

\item Choose the breakpoint that leads to the highest discrepancy 
between score functions in the two resulting subgroups. This 
difference can be measured by the linear statistic
$$
T_{j^{\ast}}^r = \sum_{i \in \mathcal{B}_{1r}} s(\hat{\theta}, y_i)
$$
where $\mathcal{B}_{1r}$ is the first of the two new subgroups that 
are defined by splitting in split point $r$ of variable $Z_{j^{\ast}}$. 
The split point is then chosen as follows:
$$
r^{\ast} = \argmin_{r} c(t_{j^{\ast}}^r,\mu_{j^{\ast}}^r,\Sigma_{j^{\ast}}^r).
$$

\item Repeat the testing and splitting procedure in each of the resulting 
subgroups until some stopping criterion is reached. This criterion 
can for example be a minimal number of observations in a node or a 
minimal p-value for the statistical tests.

This permutation test based tree algorithm is presented in 
\cite{Hothorn+Hornik+Zeileis:2006} as the ctree algorithm. 
A different framework to build a tree is provided by the MOB algorithm 
which is based on M-fluctuation tests (\citealt{Zeileis+Hothorn+Hornik:2008}).

>>>>

@@

After considering only one observation station up to now, a second step
evaluates and compares the competing methods on all 95~stations. As in the previous
section, all models are learned on the first 24~years and evaluated by the
average CRPS on the last 4~years. More specifically, the CRPS skill score
against the EMOS model is computed for the out-of-sample predictions at each
station and visualized by parallel coordinates plots with boxplots superimposed
in Figure~\ref{fig:all-crps}. Overall distributional forests have a slightly
higher improvement in CRPSS compared to the two GAMLSS which is best seen
by looking at the boxplots and also in the green line representing the results
for station Axams. The underlying parallel coordinates additionally
bring out that the prespecified GAMLSS sometimes performs rather differently
(sometimes better, sometimes worse) compared to the two data-driven models.
Finally, because some CRPSS become negative, it is also clear that the simple
EMOS performs better than the other models for a few of the stations.

@@

In the following, the tree algorithm applied in the empirical case study 
discussed in this paper is going to be explained. For notational simplicity, 
the testing and splitting procedure is described for the root node with 
observations $\{y_i\}_{i = 1,\ldots,n}$, $n \in \mathbb{N}$. 
In each child node the corresponding subset of observations depends on 
the assignment made by the foregoing split.

After fitting a distributional model $\mathcal{D}(Y, \bm{\theta})$ to 
the set of observations $\{y_i\}_{i = 1,\ldots,n}$ as explained in 
Section~\ref{sec:distfit} the resulting estimated parameter 
$\bm{\hat{\theta}} = 
(\hat{\theta}_1, \ldots, \hat{\theta}_k)$, $k \in \mathbb{N}$ 
can be plugged in the score function $s(\bm{\theta}, Y)$.
In that way a meassurement of goodness-of-fit is obtained for each
parameter $\theta_j$ and each observation $y_i$.\\
To use this information, statistical tests are employed to detect
dependencies between the score values
$$
s(\bm{\hat{\theta}}, y) = 
%\begin{pmatrix} 
%s(\bm{\hat{\theta}}, y_1) \\
%s(\bm{\hat{\theta}}, y_2) \\
%\vdots \\
%s(\bm{\hat{\theta}}, y_n)
%\end{pmatrix} =
\begin{pmatrix} 
s(\bm{\hat{\theta}}, y_1)_1 & s(\bm{\hat{\theta}}, y_1)_2 & \ldots & s(\bm{\hat{\theta}}, y_1)_k\\
s(\bm{\hat{\theta}}, y_2)_1 & s(\bm{\hat{\theta}}, y_2)_2 & \ldots & s(\bm{\hat{\theta}}, y_2)_k\\
\vdots & \vdots & \ddots & \vdots \\
s(\bm{\hat{\theta}}, y_n)_1 & s(\bm{\hat{\theta}}, y_n)_2 & \ldots & s(\bm{\hat{\theta}}, y_n)_k
\end{pmatrix}
$$
and each variable $Z_l \in \{Z_1, \ldots, Z_m\}$.  
\begin{align*}
H_0^l:  s(\bm{\hat{\theta}}, y) \qquad \bot \qquad Z_l 
\end{align*}

These hypotheses are assessed by applying permutation testing with 
the multivariate linear statistic
$$
T_{l} = \sum_{i=1}^n v_l(z_{li}) \cdot s(\bm{\hat{\theta}}, y_i).
$$
The type of the transformation function $v_l$ depends on 
the type of the split variable $Z_l$. If $Z_l$ is numeric then $v_l$ 
is simply the identity function $v_l(z_{li}) = z_{li}$. If $Z_l$ is 
a categorical variable with $H$ categories then 
$v_l(z_{li}) = e_H(z_{li}) = (\I(z_{ji} = 1), \ldots, \I(z_{li} = H))$ 
such that $v_l$ is a unit vector where the element corresponding to 
the value of $z_{li}$ is $1$. Observations with missing values are 
excluded from the sums.

With the conditional expectation $\mu_{l}$ and the covariance 
$\Sigma_{l}$ as derived by \cite{Strasser+Weber:1999} the test 
statistic can be standardized. The observed multivariate linear 
statistic $t$ is mapped into the real line by a  univariate test 
statistic $c$. In the application of this paper a quadratic form 
is chosen, such that
$$
c_{quad}(t,\mu,\Sigma) = (t-\mu)\Sigma^+(t-\mu)^{\top}
$$
where $\Sigma^+$ is the Moore-Penrose inverse of $\Sigma$. Alternatively,
the maximum of the absolute values of the standardized linear statistic
can be considered.

The smaller the p-value corresponding to the standardized test 
statistic $c(t_{l},\mu_{l},\Sigma_{l})$ is the stronger the discrepancy 
from the assumption of independence between the scores and the split 
variable $Z_l$.

After Bonferroni adjusting the p-values it has to be assessed whether
any of the resulting p-values is beneath the selected significance level. 
If so, the partitioning variable $Z_{l^\ast}$ with the lowest p-value 
is chosen as splitting variable.

As split point the breakpoint that leads to the highest discrepancy 
between score functions in the two resulting subgroups is selected. 
This difference can be measured by the linear statistic
$$
T_{l^{\ast}}^r = \sum_{i \in \mathcal{B}_{1r}} s(\bm{\hat{\theta}}, y_i)
$$
where $\mathcal{B}_{1r}$ is the first of the two new subgroups that 
are defined by splitting in split point $r$ of variable $Z_{l^{\ast}}$. 
The split point is then chosen as follows:
$$
r^{\ast} = \argmin_{r} c(t_{l^{\ast}}^r,\mu_{j^{\ast}}^r,\Sigma_{l^{\ast}}^r).
$$

Repeat the testing and splitting procedure in each of the resulting 
subgroups until some stopping criterion is reached. This criterion 
can for example be a minimal number of observations in a node or a 
minimal p-value for the statistical tests. In that way pre-pruning
is applied in order to find right-sized trees and hence 
avoid overfitting.

This permutation test based tree algorithm is presented in 
\cite{Hothorn+Hornik+Zeileis:2006} as the CTree algorithm. 
A different framework to build a tree is provided by the MOB algorithm 
which is based on M-fluctuation tests (\citealt{Zeileis+Hothorn+Hornik:2008}).
